Best Practices for Alchemical Free Energy Calculations: v\versionnumber

### 2020-01-09 paper/manuscript.tex

\newcommand{\versionnumber}{0.1}  % you should update the minor version number in preprints and major version number of submissions.
\newcommand{\githubrepository}{\url{https://github.com/michellab/alchemical-best-practices}}  %this should be the main github repository for this article
\newcommand{\expect}[1]{\left\langle{#1}\right\rangle}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Best Practices for Alchemical Free Energy Calculations: v\versionnumber}
\author[2*]{Antonia S. J. S. Mey}
\author[7]{Bryce K. Allen}
\author[1*]{John D. Chodera}
\author[2,12]{Maximilian Kuhn}
\author[2*]{Julien Michel}
\author[3*]{David L. Mobley}
\author[13]{Levi N. Naden}
\author[4]{Samarjeet Prasad}
\author[5]{Julia E. Rice}
\author[1,8]{Andrea Rizzi}
\author[2]{Jenke Scheen}
\author[6*]{Michael R. Shirts}
\author[11]{Gary Tresadern}
\author[10]{Huafeng Xu}

@@

\subsubsection{Choices unique to relative free energy calculations}

@@

\paragraph{Constraints and relative free energy calculations.}
One issue which requires particular care is the use of constraints.
Commonly, bonds involving hydrogen are constrained to a fixed length using algorithms such as SHAKE or LINCS, allowing the use of longer timesteps~\cite{}.
However, in single topology relative free energy calculations, the atoms involved might be mutated to other atom types -- for example, in a mutation of methane to methanol, one hydrogen might become an oxygen atom.
Typical molecular dynamics engines are not set up to recognize this change, or at least not to correctly include contributions to the free energy from changing constraints/constraint length, so results for a transformation would usually be erroneous.
At present the most general solution to this problem is simply to avoid the use of constraints (and thus use a smaller timestep if necessary, usually of around 1 fs) in any relative free energy calculation involving a transformation of a constrained bond.
\subsubsection{Absolute free energy calculations must handle the standard state and use restraints}
\label{sec:standardstate-restraints}
\begin{figure}
    \includegraphics[width=0.95\linewidth]{figures/fig5_thermo_cyc_abs/Figure.pdf}
    \caption{\textbf{Absolute thermodynamic cycle}}
    \label{fig:fig_absolute_thermodynamic_cycle}

@@

\subsubsection{The importance of the alchemical pathway\label{sec:important_path}}
\todo[inline, color={green!20}]{ASJM: This a bit waffly. Can we address: what do softcore potentials actually look like (equation)? Give an example of a lambda switching function(equation)? MRS: adressed these questions to some extent.  How should you chose your lambda intermediates? What is the best practice here? I have conveniently provided a figure to help with the discussion.}
\todo[inline, color={red!40}]{LNN: write common principles alchemical path choice. MRS: mostly wrote this, LNN can comment on it,}
\todo[inline, color={red!40}]{AR: write absolute-specific section on  alchemical path choice}
\todo[inline, color={red!40}]{BA: write relative-specific section on  alchemical path choice}

@@

For many energy terms, \textit{as long as a repulsive core remains on}, this is a very good approach.  For example, it can be shown that if van der Waals repulsions are left on, then the linear approach is very nearly the optimal path  possible for changing, removing, or inserting the electrostatic energy terms, with the path being within about \%10--20 of the minimum possible uncertainty~\cite{naden2015linear} for a fixed amount of simulation, as well as being nearly optimally efficient for van der Waals attractive terms with repulsion terms turned on~\cite{naden2014linear}. Although we are not aware of any quantitative tests for dipolar or higher multipole terms, theoretically it should behave equally well for those systems.

@@

What we need instead is some sort of function that smoothly gets rid of this infinity.  A large number of scheme have been tried~\cite{beutler1994avoiding, zacharias1994separationshifted, blondel2004ensemble, pham2011identifying, pham2012optimal, naden2014linear, beutler1994avoiding, donnini2005molecular}, but the most common strategy that appears to be the best practices is to use a "soft-core" potential, of the form:

@@

\section*{Author Contributions}
%%%%%%%%%%%%%%%%
% This section mustt describe the actual contributions of
% author. Since this is an electronic-only journal, there is
% no length limit when you describe the authors' contributions,
% so we recommend describing what they actually did rather than
% simply categorizing them in a small number of
% predefined roles as might be done in other journals.
%
% See the policies ``Policies on Authorship'' section of https://livecoms.github.io
% for more information on deciding on authorship and author order.
%%%%%%%%%%%%%%%%
\todo[inline, color={green!20}]{ASJSM: @everyone, can you check your contribution is acknowledge and if not add the appropriate information. }
(Explain the contributions of the different authors here)
ASJSM: Coordinated the document, contributed to most sections, and co-designed figures ~\ref{} and created figures \ref{fig:fig_what_is_lambda},~\ref{fig:fig1_what_is_alchemy},~\ref{fig:overlap} and replotted~\ref{fig:automatic-equilibration-detection} and improved ~\ref{fig:fig_types_of_networks}.
JS: created figures \ref{fig:fig_what_is_alchemy},~\ref{fig:fig_binding_thermodynamic_cycle},~\ref{fig:fig_topology},~\ref{fig:fig_mcss},~\ref{fig:fig_types_of_networks},~\ref{fig:fig_absolute_thermodynamic_cycle},~\ref{fig:scatterplot_analysis}, the data representation section, the checklist and contributed to general formatting discussions and editing.
MK: Contributed to the Data Analysis section, provided the data for figure~\ref{fig:overlap} and helped edit the paper.
DLM: Contributed to the outline, drafted some of the sections, gave ideas on figures, and helped edit the paper.
GT: Contributed to introduction and drug discovery sections, and helped edit the paper.
AR: Created figure~\ref{fig:freeenergytrajectories}, contributed to section \ref{sec:step2}, and helped edit the paper.
MRS: Helped create figure~\ref{fig:fig_what_is_lambda}, wrote sections describing choices for alchemical pathways and analysis for free energy calculations.
LNN: Helped write the simulation length, stopping conditions, and information saving section.
BA: \todo[inline, color={red!20}]{@Bryce insert info here}
JM: \todo[inline, color={red!20}]{@Julien insert info here}
JDC: \todo[inline, color={red!20}]{@JDC insert info here}
HX: \todo[inline, color={red!20}]{@Huafeng insert info here}
% We suggest you preserve this comment:
For a more detailed description of author contributions,
see the GitHub issue tracking and changelog at \githubrepository.

>>>>

\newcommand{\versionnumber}{0.1}  % you should update the minor version number in preprints and major version number of submissions.
\newcommand{\githubrepository}{\url{https://github.com/michellab/alchemical-best-practices}}  %this should be the main github repository for this article
\newcommand{\expect}[1]{\left\langle{#1}\right\rangle}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Best Practices for Alchemical Free Energy Calculations: v\versionnumber}
\author[2*]{Antonia S. J. S. Mey}
\author[7]{Bryce K. Allen}
\author[1*]{John D. Chodera}
\author[2,12]{Maximilian Kuhn}
\author[2*]{Julien Michel}
\author[3*]{David L. Mobley}
\author[13]{Levi N. Naden}
\author[4]{Samarjeet Prasad}
\author[5]{Julia E. Rice}
\author[1,8]{Andrea Rizzi}
\author[2]{Jenke Scheen}
\author[6*]{Michael R. Shirts}
\author[11]{Gary Tresadern}
\author[7]{Huafeng Xu}

@@

\subsubsection{Choices unique to relative free energy calculations}
\label{sec:relative-fe-protocol}

@@

Methods in experimental design have been applied to the construction of the perturbation maps.  Yang et al.~\cite{yang2019optimal} optimized the perturbation map by selecting a fixed number of calculations from the pairwise perturbations so that the resulting set of calculations minimize the total variance.  Xu~\cite{xu2019diffnet} optimized the perturbation map by allocating different amounts of simulation time to different pairwise perturbations so as to minimize the total variance, given the total simulation time of all the perturbation calculations.  Both approaches lead to substantial reduction in the statistical error of the estimated free energies.  
%
\paragraph{Constraints and relative free energy calculations.}
One issue which requires particular care is the use of constraints.
Commonly, bonds involving hydrogen are constrained to a fixed length using algorithms such as SHAKE or LINCS, allowing the use of longer timesteps~\cite{}.
However, in single topology relative free energy calculations, the atoms involved might be mutated to other atom types -- for example, in a mutation of methane to methanol, one hydrogen might become an oxygen atom.
Typical molecular dynamics engines are not set up to recognize this change, or at least not to correctly include contributions to the free energy from changing constraints/constraint length, so results for a transformation would usually be erroneous.
At present the most general solution to this problem is simply to avoid the use of constraints (and thus use a smaller timestep if necessary, usually of around 1 fs) in any relative free energy calculation involving a transformation of a constrained bond.
%
\subsubsection{Absolute free energy calculations must handle the standard state and use restraints}
\label{sec:standardstate-restraints}
%
\begin{figure}
    \includegraphics[width=0.95\linewidth]{figures/fig5_thermo_cyc_abs/Figure.pdf}
    \caption{\textbf{Absolute thermodynamic cycle}}
    \label{fig:fig_absolute_thermodynamic_cycle}
\end{figure}

@@

\subsubsection{The importance of the alchemical pathway\label{sec:important_path}}
\todo[inline, color={green!20}]{ASJM: This a bit waffly. Can we address: what do softcore potentials actually look like (equation)? Give an example of a lambda switching function(equation)? MRS: adressed these questions to some extent.  How should you chose your lambda intermediates? What is the best practice here? I have conveniently provided a figure to help with the discussion.}
\todo[inline, color={red!40}]{LNN: write common principles alchemical path choice. MRS: mostly wrote this, LNN can comment on it. LNN: LGTM}
\todo[inline, color={red!40}]{AR: write absolute-specific section on  alchemical path choice}
\todo[inline, color={red!40}]{BA: write relative-specific section on  alchemical path choice}

@@

For many energy terms, \textit{as long as a repulsive core remains on}, this is a very good approach.  For example, it can be shown that if van der Waals repulsions are left on, then the linear approach is very nearly the optimal path  possible for changing, removing, or inserting the electrostatic energy terms, with the path being within about 10--20\% of the minimum possible uncertainty~\cite{naden2015linear} for a fixed amount of simulation, as well as being nearly optimally efficient for van der Waals attractive terms with repulsion terms turned on~\cite{naden2014linear}. Although we are not aware of any quantitative tests for dipolar or higher multipole terms, theoretically it should behave equally well for those systems.

@@

What we need instead is some sort of function that smoothly gets rid of this infinity.  A large number of scheme have been tried~\cite{beutler1994avoiding, zacharias1994separationshifted, blondel2004ensemble, pham2011identifying, pham2012optimal, naden2014linear, donnini2005molecular}, but the most common strategy that appears to be the best practices is to use a "soft-core" potential, of the form:

@@

\section*{Author Contributions}
%%%%%%%%%%%%%%%%
% This section mustt describe the actual contributions of
% author. Since this is an electronic-only journal, there is
% no length limit when you describe the authors' contributions,
% so we recommend describing what they actually did rather than
% simply categorizing them in a small number of
% predefined roles as might be done in other journals.
%
% See the policies ``Policies on Authorship'' section of https://livecoms.github.io
% for more information on deciding on authorship and author order.
%%%%%%%%%%%%%%%%
\todo[inline, color={green!20}]{ASJSM: @everyone, can you check your contribution is acknowledge and if not add the appropriate information. }
(Explain the contributions of the different authors here)
ASJSM: Coordinated the document, contributed to most sections, and co-designed figures ~\ref{} and created figures \ref{fig:fig_what_is_lambda},~\ref{fig:fig1_what_is_alchemy},~\ref{fig:overlap} and replotted~\ref{fig:automatic-equilibration-detection} and improved ~\ref{fig:fig_types_of_networks}.
JS: created figures \ref{fig:fig_what_is_alchemy},~\ref{fig:fig_binding_thermodynamic_cycle},~\ref{fig:fig_topology},~\ref{fig:fig_mcss},~\ref{fig:fig_types_of_networks},~\ref{fig:fig_absolute_thermodynamic_cycle},~\ref{fig:scatterplot_analysis}, the data representation section, the checklist and contributed to general formatting discussions and editing.
MK: Contributed to the Data Analysis section, provided the data for figure~\ref{fig:overlap} and helped edit the paper.
DLM: Contributed to the outline, drafted some of the sections, gave ideas on figures, and helped edit the paper.
GT: Contributed to introduction and drug discovery sections, and helped edit the paper.
AR: Created figure~\ref{fig:freeenergytrajectories}, contributed to section \ref{sec:step2}, and helped edit the paper.
MRS: Helped create figure~\ref{fig:fig_what_is_lambda}, wrote sections describing choices for alchemical pathways and analysis for free energy calculations.
LNN: Helped write the simulation length, stopping conditions, and information saving section.
BA: \todo[inline, color={red!20}]{@Bryce insert info here}
JM: Contributed to sections~\ref{subsec:reproducible},~\ref{sec:step1},~\ref{} 4.2, 5.3, 6, 8.3.1, 8.3.2\todo[inline, color={red!20}]{@Julien insert info here}
JDC: \todo[inline, color={red!20}]{@JDC insert info here}
HX: Contributed section~\ref{subsec:accuracy} and to section~\ref{sec:relative-fe-protocol}.

% We suggest you preserve this comment:
For a more detailed description of author contributions,
see the GitHub issue tracking and changelog at \githubrepository.
